Categories of Optics (Draft)

### 2018-08-19 paper.tex
Delete a lot of old commented out stuff

In the Haskell \lenslib{} library, optics are written in a form known as the \emph{profunctor encoding}, which at first glance is completely different to that given in Section~\ref{sec:optics}. As this was being written, Bartosz Milewski~\cite{ProfunctorOpticsPost} independently described the isomorphism between some forms of optic and their profunctor encoding. In Section~\ref{sec:profunctor-optics} we review this equivalence and verify that the folklore profunctor optic laws are equivalent to lawfulness as defined here.

@@

%\todo{Past work:
%  \begin{itemize}
%  \item The Haskell lens people: edwardk, shachaf, xplat, roconnor...
%  \item The Doubles paper
%  \item Bartosz's blog post which was written about the same time I started this. (Only tensor-like optics, no discussion of laws, confusion about enrichment(?))
%  \item The profunctor optics paper out of Oxford. (No generality, nothing categorical, no discussion of laws)
%  \item Very old Dialectica category stuff
%  \end{itemize}
%}

@@

  Note that because $\iota(S, S') = (S, S')$, required equations for $\iota$ to be a monoidal functor hold by definition (although we don't yet know that $\Optic_\C$ is monoidal). The pentagon and triangle equations then hold in $\Optic_\C$, as they are the image of the same diagrams in $\C \times \C^\op$ under $\iota$. The only remaining thing to verify is that these structure maps are natural in $\Optic_\C$, but this follows from the previous proposition.

%  Let us check that $\alpha_{(R, R'), (S, S'), (T, T')}$ is natural in $(R, R')$. Suppose $\rep{l}{r} : (Q, Q') \hto (R, R')$ is an optic with residual $M$.
%  \begin{align*}
%    &\alpha_{(R, R'), (S, S'), (T, T')} \circ (f \otimes (S, S')) \otimes (T, T') \\
%    &= \iota(\alpha_{R,S,T}, \alpha_{R',S',T'}^{-1})  \circ (\rep{l}{r} \otimes (S, S')) \otimes (T, T') \\
%    &= \rep{\lambda_{R \otimes (S \otimes T)}\alpha_{R,S,T}}{\alpha_{R',S',T'}^{-1} \lambda^{-1}_{(R' \otimes S') \otimes T'}}  \circ \rep{(l \otimes S) \otimes T}{(r \otimes S') \otimes T'} \\
%    &= \rep{(M \otimes \lambda_{R \otimes (S \otimes T)}\alpha_{R,S,T})((l \otimes S) \otimes T)}{((r \otimes S') \otimes T') (M \otimes \alpha_{R',S',T'}^{-1} \lambda^{-1}_{(R' \otimes S') \otimes T'} )} \\
%    &=\rep{(M \otimes \alpha_{R,S,T})((l \otimes S) \otimes T)}{((r \otimes S') \otimes T') (M \otimes \alpha_{R',S',T'}^{-1} )} \\
%    &= \rep{(l \otimes (S \otimes T))\alpha_{Q,S,T}}{\alpha_{Q',S',T'}^{-1}(r \otimes (S' \otimes T'))} \\
%    &= \rep{(I \otimes l \otimes (S \otimes T)) \lambda_{Q \otimes (S \otimes T)}\alpha_{Q,S,T}}{(I \otimes r \otimes (S' \otimes T')) \alpha_{Q',S',T'}^{-1} \lambda^{-1}_{Q' \otimes (S' \otimes T')}} \\
%    &= \rep{l \otimes (S \otimes T)}{r \otimes (S' \otimes T')} \circ \rep{\lambda_{Q \otimes (S \otimes T)}\alpha_{Q,S,T}}{\alpha_{Q',S',T'}^{-1} \lambda^{-1}_{Q' \otimes (S' \otimes T')} } \\
%    &= \rep{l}{r} \otimes ((S, S') \otimes (T, T')) \circ \iota(\alpha_{Q,S,T}, \alpha_{Q',S',T'}^{-1}) \\
%    &= f \otimes ((S, S') \otimes (T, T')) \circ \alpha_{(Q, Q'), (S, S'), (T, T')}
%  \end{align*}
%
%Naturality in the other variables is similar, as is the naturality of the unitors.
%  \todo{convert this to a string diagram?}
\end{proof}

@@

% \todo{This doesn't give us the structure of a `framing', we can only lift vertical arrows to horizontal arrows facing one direction.}

% \begin{theorem}
%   $\Optic_\C$ is symmetric monoidal, where $(S, S') \otimes (T, T') = (S \otimes T, S' \otimes T')$, the unit is $(I, I)$, and the action on morphisms induced by
%     &\C(S, M \otimes A) \times \C(M \otimes A', S') \times \C(T, N \otimes B) \times \C(N \otimes B', T')\\
%     \to \quad&\C(S \otimes T, M \otimes A \otimes N \otimes B) \times \C(M \otimes A' \otimes N \otimes B', S' \otimes T') && \text{(action of $\otimes$ on morphisms in $\C$)}\\
%     \to \quad&\C(S \otimes T, M \otimes N \otimes A \otimes B) \times \C(M \otimes N \otimes A' \otimes B', S' \otimes T') && \text{($\otimes$ in $\C$ is symmetric)}\\
%     \to \quad&\int^{M \in \C} \C(S \otimes T, M \otimes A \otimes B) \times \C(M \otimes A' \otimes B', S' \otimes T') && \text{(inclusion into coend)}
% \end{theorem}
% \begin{proof}
%   Equationally, suppose we have representatives $(l, r) : (S \to M \otimes A, M \otimes A' \to S')$ and $(l', r') : (T \to N \otimes B, N \otimes B' \to T')$. Then:
%     \langle l, r \rangle \otimes \langle l', r' \rangle &:= \langle (M \otimes s_{A,N} \otimes B)(l \otimes l'), (r \otimes r')(M \otimes s_{A',N} \otimes B') \rangle
%   This is functorial, as suggested by the equivalence of the following string diagrams:
%   \todo{todo}
%%   For future convenience, note that
%%   \begin{align*}
%%     \langle l, r \rangle \otimes (T, T') &= \langle (M \otimes s_{A,I} \otimes T)(l \otimes \lambda_T), (r \otimes \lambda^{-1}_{T'})(M \otimes s_{A',I}\otimes T') \rangle \\
%%     &= \langle (\lambda^{-1}_M \otimes A \otimes T)(M \otimes s_{A,I} \otimes T)(l \otimes \lambda_T), (r \otimes \lambda^{-1}_{T'})(M \otimes s_{A',I}\otimes T')(\lambda_M \otimes A' \otimes T') \rangle \\
%%     &= \langle l \otimes T, r \otimes T' \rangle
%%   \end{align*}
%%   Now we check functoriality. Suppose we have:
%%   \begin{align*}
%%     \langle l_1, r_1 \rangle : (S_1, S_1') &\hto (S_2, S_2') \\
%%     \langle l_2, r_2 \rangle : (S_2, S_2') &\hto (S_3, S_3') \\
%%     \langle p_1, q_1 \rangle : (T_1, T_1') &\hto (T_2, T_2') \\
%%     \langle p_2, q_2 \rangle : (T_2, T_2') &\hto (T_3, T_3')
%%   \end{align*}
%%   with residuals $M_1, M_2, N_1, N_2$ respectively. Then: \todo {(This is a mess. String diagram?)}
%%   \begin{align*}
%%     &(\langle l_2, r_2 \rangle \otimes \langle p_2, q_2 \rangle) \circ (\langle l_1, r_1 \rangle \otimes \langle p_1, q_1 \rangle) \\
%%     = \; &\langle (M_2 \otimes s_{S_3,N_2} \otimes T_3)(l_2 \otimes p_2), (r_2 \otimes q_2)(M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
%%     &\circ \langle (M_1 \otimes s_{S_2,N_1} \otimes T_2)(l_1 \otimes p_1), (r_1 \otimes q_1)(M_1 \otimes s_{S_2',N_1} \otimes T_2') \rangle \\
%%     = \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes N_1 \otimes l_2 \otimes p_2)(M_1 \otimes s_{S_2,N_1} \otimes T_2)(l_1 \otimes p_1), \\
%%     &\;(r_1 \otimes q_1)(M_1 \otimes s_{S_2',N_1} \otimes T_2')(M_1 \otimes N_1 \otimes r_2 \otimes q_2)(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
%%     = \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes s_{(M_2 \otimes S_3),N_1} \otimes T_3)(M_1 \otimes  l_2 \otimes N_1 \otimes p_2)(l_1 \otimes p_1), \\
%%     &\;(r_1 \otimes q_1)(M_1 \otimes r_2 \otimes N_1 \otimes q_2)(M_1 \otimes s_{(M_2 \otimes S_3'),N_1} \otimes T_3')(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
%%     = \; & \langle (M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3,N_2} \otimes T_3)(M_1 \otimes s_{(M_2 \otimes S_3),N_1} \otimes T_3)((M_1 \otimes l_2)l_1\otimes (N_1 \otimes p_2)p_1), \\
%%     &\;(r_1(M_1 \otimes r_2) \otimes q_1(N_1 \otimes q_2))(M_1 \otimes s_{(M_2 \otimes S_3'),N_1} \otimes T_3')(M_1 \otimes N_1 \otimes M_2 \otimes s_{S_3',N_2} \otimes T_3') \rangle \\
%%     = \; & \text{\todo{(Add a $s_{M_2, N_1}$ to either side of the coend, then appeal to coherence)}} \\
%%     = \; &\langle (M_1 \otimes M_2 \otimes s_{(N_1 \otimes N_2),S_3} \otimes T_3)((M_1 \otimes l_2)l_1\otimes (N_1 \otimes p_2)p_1) ,  \\
%%     &\;(r_1(M_1 \otimes r_2) \otimes q_1(N_1 \otimes q_2))(M_1 \otimes M_2 \otimes s_{(N_1 \otimes N_2),S'_3} \otimes T'_3) \rangle \\
%%     = \; &\langle (M_1 \otimes l_2)l_1, r_1(M_1 \otimes r_2) \rangle \otimes \langle (N_1 \otimes p_2)p_1, q_1(N_1 \otimes q_2) \rangle \\
%%     = \; &(\langle l_2, r_2 \rangle  \circ \langle l_1, r_1 \rangle) \otimes (\langle p_2, q_2 \rangle \circ \langle p_1, q_1 \rangle)
%%   \end{align*}
%
%   The structure morphisms are all lifted from the structure morphisms in $\C$:
%     \alpha_{(R, R'), (S, S'), (T, T')} &:= \iota(\alpha_{R,S,T}, \alpha_{R',S',T'}^{-1}) \\
%     \lambda_{(S, S')} &:= \iota(\lambda_{(S, S')}, \lambda_{(S, S')}^{-1}) \\
%     \rho_{(S, S')} &:= (\rho_{(S, S')}, \rho_{(S, S')}^{-1}) \\
%     s_{(S, S'), (T, T')} &:= \iota(s_{S, T}, s_{T', S'})
%
%   Note that because $\iota(S, S') = (S, S')$, required equations for $\iota$ to be a monoidal functor hold by definition (although we don't yet know that $\Optic_\C$ is monoidal). The pentagon and triangle axioms then hold in $\Optic_\C$, as they are the image of the same axioms in $\C \times \C^\op$ under $\iota$. The only remaining thing to verify is that these structure maps are natural in $\Optic_\C$.
%
%   Let us check that $\alpha_{(R, R'), (S, S'), (T, T')}$ is natural in $(R, R')$. Suppose $f : (Q, Q') \hto (R, R')$ has representative $(l, r) : (Q \to M \otimes R, M \otimes R' \to Q')$, then:
%     &\alpha_{(R, R'), (S, S'), (T, T')} \circ (f \otimes (S, S')) \otimes (T, T') \\
%     &= \iota(\alpha_{R,S,T}, \alpha_{R',S',T'}^{-1})  \circ (\langle l, r \rangle \otimes (S, S')) \otimes (T, T') \\
%     &= \langle\lambda_{R \otimes (S \otimes T)}\alpha_{R,S,T}, \alpha_{R',S',T'}^{-1} \lambda^{-1}_{(R' \otimes S') \otimes T'} \rangle  \circ \left\langle (l \otimes S) \otimes T, (r \otimes S') \otimes T' \right\rangle \\
%     &=\left\langle (M \otimes \lambda_{R \otimes (S \otimes T)}\alpha_{R,S,T})((l \otimes S) \otimes T), ((r \otimes S') \otimes T') (M \otimes \alpha_{R',S',T'}^{-1} \lambda^{-1}_{(R' \otimes S') \otimes T'} ) \right\rangle \\
%     &=\left\langle (M \otimes \alpha_{R,S,T})((l \otimes S) \otimes T), ((r \otimes S') \otimes T') (M \otimes \alpha_{R',S',T'}^{-1} )\right\rangle \\
%     &=\left\langle (l \otimes (S \otimes T))\alpha_{Q,S,T}, \alpha_{Q',S',T'}^{-1}(r \otimes (S' \otimes T')) \right\rangle \\
%     &= \langle (I \otimes l \otimes (S \otimes T)) \lambda_{Q \otimes (S \otimes T)}\alpha_{Q,S,T}, (I \times r \otimes (S' \otimes T')) \alpha_{Q',S',T'}^{-1} \lambda^{-1}_{Q' \otimes (S' \otimes T')}\rangle \\
%     &= \langle l \otimes (S \otimes T), r \otimes (S' \otimes T')\rangle \circ \left\langle \lambda_{Q \otimes (S \otimes T)}\alpha_{Q,S,T}, \alpha_{Q',S',T'}^{-1} \lambda^{-1}_{Q' \otimes (S' \otimes T')} \right\rangle \\
%     &= \langle l, r \rangle \otimes ((S, S') \otimes (T, T')) \circ \iota(\alpha_{Q,S,T}, \alpha_{Q',S',T'}^{-1}) \\
%     &= f \otimes ((S, S') \otimes (T, T')) \circ \alpha_{(Q, Q'), (S, S'), (T, T')}
%   \todo{There is some cheating going on here, composition actually adds an extra $\alpha$, but that wouldn't get in the way.}
% \end{proof}

@@

% \begin{proposition}
%   Any isomorphism $p : (S, S') \hto (A, A')$ in $\Optic_\C$ is of the form $\iota(f, g)$ for two isomorphisms $S \to A$ and $A' \to S'$ in $\C$.
% \end{proposition}
% \begin{proof}
%   \todo{This might not actually be true..}
%\end{proof}

@@

\begin{definition}[Compare {\cite[Definition 5.1]{CoherenceForLenses}}]
  A \emph{teleological category is} a symmetric monoidal category $(\T, \teletimes, I)$, equipped with:
  \begin{itemize}
  \item A symmetric monoidal subcategory $\T_d$ of \emph{dualisable morphisms} containing all the objects of $\T$, with an involutive symmetric monoidal functor ${(-)}^* : \T_d \to \T_d^\op$, where---not finding a standard symbol for such a thing---we mean $\T_d^\op$ to be the category with both the direction of the arrows \emph{and} the order of the tensor flipped: ${(A \otimes B)}^* \cong B^* \otimes A^*$. Note that there is therefore also a canonical isomorphism $\phi : I \cong I^*$
  \item A symmetric monoidal extranatural family of morphisms $\varepsilon_X : X \otimes X^* \to I$, called \emph{counits}, natural with respect to the morphisms in $\T_d$.
  \end{itemize}
\end{definition}
Unpacking the definition, $\varepsilon$ being a symmetric monoidal extranatural transformation amounts to the following diagrams in $\T$ commuting:
\[
  \begin{tikzcd}
    X \teletimes Y^* \ar[r, "f \teletimes Y^*"]  \ar[d, "X \teletimes f^*", swap] & Y \teletimes Y^* \ar[d, "\varepsilon_Y"] \\
    X \teletimes X^* \ar[r, "\varepsilon_X", swap] & I
  \end{tikzcd} \hspace{1cm}
  \begin{tikzcd}
    X^* \teletimes X \ar[r, "s"]  \ar[dr, "\varepsilon_{X^*}", swap] & X \teletimes X^* \ar[d, "\varepsilon_X"] \\
    & I
  \end{tikzcd}\]
\[
  \begin{tikzcd}[column sep = large]
    X \teletimes Y \teletimes Y^* \teletimes X^* \ar[r, "X \teletimes \varepsilon_Y \teletimes X^*"]  \ar[dr, "\varepsilon_{X \teletimes Y}", swap] & X \teletimes X^* \ar[d, "\varepsilon_X"] \\
    & I
  \end{tikzcd} \hspace{1cm}
  \begin{tikzcd}
    I \teletimes I^* \ar[r,"I \teletimes \phi"] \ar[dr, swap, "\varepsilon_I"] & I \ar[d, equal] \\
    & I
  \end{tikzcd}
\]
where $f : X \to Y$ is in $\T_d$.

@@

% \begin{remark}
%   Compact closed categories very nearly form examples of teleological categories. In a compact closed category, there are canonical isomorphisms $(X \otimes Y)^* \cong Y^* \otimes X^*$. In a teleological category, however, duality does not reverse the order of the tensor. \todo{If $\C$ is strict, as it is here, does this matter?}
% \end{remark}

@@

%There are few surprises in the following definitions.

@@

      %       \todo{They stress about the empty set, have I avoided that issue?}

@@

      %       \begin{proposition}
      %       Suppose $\langle l, r \rangle : (S, S') \hto (A, A')$ has residual $M$. Then
      %       \begin{align*}
      %       \langle l, r \rangle = (\varepsilon_{(M, I)} \otimes (A, A'))\iota(l, r).
      %     \end{align*}
      %       \end{proposition}
      %       \begin{proof}
      %       First note that because $\C$ is strict monoidal, the counit $\varepsilon_{(M, I)} : (M \otimes I, M \otimes I) = (M, M) \hto (I, I)$ is equal to the connector $c_M : (M, M) \hto (I, I)$.
      %
      %       The claim then holds, as any optic
      %       \begin{center}
      %       \input{diagrams/generic-optic-noline.tikz}
      %       \end{center}
      %       is equal to the composite
      %       \begin{center}
      %       \input{diagrams/optic-decomposed.tikz}
      %       \end{center}
      %       where we have drawn the normally invisible strings corresponding to the unit $I$.
      %       \end{proof}

@@

In this section we make a different definition of lawful that at first seems strange, but which in the case of lenses corresponds \emph{exactly} to the three concrete lens laws with no additional assumptions on $\C$ required. As further justification for this definition, in Section~\ref{sec:profunctor-optics} we will see an interpretation of (unlawful) optics as maps between certain comonoid objects. Lawfulness corresponds exactly to this map being a comonoid homomorphism.

@@

\begin{proposition}\label{prop:lawful-lens-laws}
  A concrete lens given by $\fget$ and $\fput$ is lawful (in our sense) iff it obeys the three concrete lens laws.
\end{proposition}
\begin{proof}
  We begin by giving $\Twoptic_\times(S, A)$ the same treatment as $\Optic_\times(S, A)$. Using Yoneda reduction and the universal property of the product as we did for lenses themselves, we have:
  \begin{align*}
  \Twoptic_\times(S, A)
  &= \int^{M_1, M_2 \in \C} \C(S, M_1 \times A) \times \C(M_1 \times A, M_2 \times A) \times \C(M_2 \times A, S) \\
  &\cong \int^{M_1, M_2 \in \C} \C(S, M_1) \times \C(S, A) \times \C(M_1 \times A, M_2 \times A) \times \C(M_2 \times A, S) \\
  &\cong \int^{M_2 \in \C} \C(S, A) \times \C(S \times A, M_2 \times A) \times \C(M_2 \times A, S) \\
  &\cong \int^{M_2 \in \C} \C(S, A) \times \C(S \times A, M_2) \times \C(S \times A, A) \times \C(M_2 \times A, S) \\
  &\cong \C(S, A) \times \C(S \times A, A) \times \C(S \times A \times A, S)
  \end{align*}
  Call this last set $\conctwice_\times(S, A)$. Written equationally, the isomorphism $\Phi : \Twoptic_\times(S, A) \to \conctwice_\times(S, A)$ is given by:
  \begin{align*}
    \Phi(\repthree{l}{c}{r}) = (&\pi_2 l, \\
                                   &\pi_2 c (\pi_1 l \times A), \\
                                   &r(\pi_1 c (\pi_1 l \times A) \times A))
  \end{align*}
  Now suppose we are given a lens $p$ that corresponds concretely to $(\fget, \fput)$, so $p = \rep{[\id_S, \fget]}{\fput}$. Evaluating $\outside$ on this gives:
  \begin{align*}
    \outside(\rep{[\id_S, \fget]}{\fput}) = \fput [\id_S, \fget]
  \end{align*}
  so requiring $\outside(p) = \id_S$ is precisely the $\fget\fput$ law.

@@

This characterisation of setters is maybe a little odd, in that we have ended up with a function of $\Set$s, rather than a description internal to $\C$. If we modify our definition of $\Setter$, we do get an internal characterisation. Suppose $\C$ is cartesian closed and let $\Strong_\C$ be the category of \emph{strong functors} on $\C$. %\todo{I define strong functor here because I also want to talk about strong monads later, should this definition wait for that?}

\begin{definition}[\cite{StrongFunctors}]\label{def:strong-functor}
  A \emph{(left) strong functor} is a functor $F : \C \to \C$ equipped with a natural transformation called the \emph{strength}:
  \begin{align*}
    \theta_{A,B} : A \otimes F B \to F(A \otimes B)
  \end{align*}
  such that the strength commutes with the unitor:
  \[
    \begin{tikzcd}
      I \otimes F A \ar[r, "\theta_{1,A}"] \ar[d, "\cong" left]  & F(I \otimes A) \ar[d, "\cong" right] \\
      F A \ar[r, equals] & F A
    \end{tikzcd}
  \]
  and with associativity:
  \[
    \begin{tikzcd}
      (A \otimes B) \otimes F C \ar[rr, "\theta_{A \otimes B, C}"] \ar[d, "\alpha_{A,B,FC}" left]  && F((A \otimes B) \otimes C) \ar[d, "F\alpha_{A,B,FC}" right] \\
      A \otimes (B \otimes F C) \ar[r, "A \otimes \theta_{B,C}" below] & A \otimes F(B \otimes C) \ar[r, "\theta_{A, B\otimes C}" below] & F(A \otimes (B \otimes C))
    \end{tikzcd}
  \]
  A \emph{strong natural transformation} $\tau : (F,\theta) \Rightarrow (G,\theta')$ is a natural transformation that respects the strengths. There is an evident category $\Strong(\C)$ of strong endofunctors and strong natural transformations, and a forgetful functor $U : \Strong(\C) \to [\C, \C]$.
%  \[
%    \begin{tikzcd}
%      A \otimes F B \ar[r, "\theta_{A,B}"] \ar[d, "A \otimes \tau_B" left]  & F(A \otimes B) \ar[d, "\tau_{A \otimes B}" right] \\
%      A \otimes G B \ar[r, "\theta'_{A, B}"]& G(A \otimes B)
%    \end{tikzcd}
%  \]
%
\end{definition}
Then, again, $\Strong_\C$ acts on $\C$ by evaluation. We leave it to the reader to verify there is a natural isomorphism \[\C(S, FA) \cong \Strong_\C(\homC(A, -) \times S, F),\] which we can use to describe optics for this action as elements of $\C(\homC(A, A'), \homC(S,S'))$.
%The use of strong functors here is due to the correspondence between tensorial strengths and ``$\C$-enrichments'' on a functor.

@@

%\begin{remark}\label{remark:four-parameters}
%We note here that Haskell lenses
%\end{remark}

%\todo{What about the pointwise tensor of an entire functor?}

@@

\item \emph{Grates}~\cite{GratesPost} are optics for the contravariant action of a monoidal closed category $\C$ on itself by $X \cdot A \mapsto \homC(X, A)$. Concretely these correspond to morphisms \[ \C(\homC(\homC(S, A), A'), S'). \]

%\item \emph{Adjunction Lenses}: For a category with finite products $\C$ and a monoidal adjunction $F \dashv U$, with $F : \C \to \D$ and $U : \D \to \C$. If $U$ is strong monoidal then $\D$ acts on $\C$ by $UD \act A := D \times A$; as an example consider the free-forgetful adjunction between sets and monoids. Concrete optics for this action are given by elements of
%\[ \C(S, A) \times \C(UFS \times A', S') \]
%\todo{This example is a bit dumb, delete?}
\end{itemize}

% \subsection{Equivalence of Optics}
%
% One issue with the definition of optic given in the last section is that it could be difficult to determine when two optics $\rep{l}{r}, \langle l', r' \rangle : S \hto A$
% are equal. For some choices of $\M$, however, we can reduce a zigzag of relations to a single one.
%
% \todo{This may work in general with some huge number of conditions: something like $-A$ reflects monomorphisms and preserves binary coproducts, all subobjects in $\M$ have complements.}
%
% \begin{proposition}
%   Suppose all epimorphisms split in $\C$, and that $- \times A : \C \to \C$ reflects epis. Then two lenses $\langle l, r \rangle$ and $\langle l', r' \rangle :  S \hto A$ with residuals $M$ and $N$ respectively are equal iff there exists an morphism $\phi : M \to N$ such that $(\phi \times A)l = l'$ and $r = r' (\phi \times A)$.
% \end{proposition}
% \begin{proof}
%   The backward direction is clear: the morphism $\phi$ is a witness that $\langle l, r \rangle \rightsquigarrow \langle l', r' \rangle$, so the two optics are equal.
%
%%   note that $p$ and $q$ must have identical $\fget$ and $\fput$ functions:
%%   \begin{align*}
%%     \fget &= \pi_2 l = \pi_2 l' \\
%%     \fput &= r (\pi_1 l \times A) = r' (\pi_1 l' \times A)
%%   \end{align*}
%   For the forward direction, note that the map $(\pi_1 l \times A)$ is a retraction: it has the section $(r, \pi_2)$:
%   \begin{align*}
%     (\pi_1 l \times A)(r, \pi_2) = (\pi_1 l r, \pi_2) = (\pi_1, \pi_2) = \id_{S \times A}
%   \end{align*}
%   Now by assumption, $\pi_1 l : S \to M$ is an epi, so has a retraction $f : M \to S$. The claim is that $\phi = \pi_1 l' f$ is the required morphism $M \to N$.
%
%   The situation is summarised in the following diagram, which other than the dashed line, we know commutes.
%   \[
%     \begin{tikzcd}
%       M \times A \ar[r, "{f \times A}", dashed, bend left = 50] \ar[d, "r", swap] & S \times A \ar[d, "\fput"] \ar[r, "{\pi_1 l' \times A}"] \ar[l, "{\pi_1 l \times A}", swap] & N \times A \ar[d, "r'"] \\
%       S \ar[r, equal] \ar[d, "l", swap] & S \ar[d, "{[\id_S, \fget]}"] \ar[r, equal] & S \ar[d, "l'"] \\
%       M \times A & S \times A \ar[r] \ar[l] & N \times A
%     \end{tikzcd}
%   \]
%   We can now calculate:
%   \begin{align*}
%     r = r(\pi_1 l f \times A) = \fput (f \times A) = r' (\pi_1 l' f \times A) = r'(\phi \times A)
%   \end{align*}
%   as required. The equation relating $l$ and $l'$ follows immediately as they are the inverses of $r$ and $r'$.
% \end{proof}
%
% The conditions of the above Proposition hold in $\Set$, so long as $A$ is not empty.

% Old attempt:
% \begin{definition}

>>>>

In the Haskell \lenslib{} library, optics are written in a form known as the \emph{profunctor encoding}, which at first glance is completely different to that given in Section~\ref{sec:optics}. As this was being written, Bartosz Milewski~\cite{ProfunctorOpticsPost} independently described the isomorphism between optics and their profunctor encoding. In Section~\ref{sec:profunctor-optics} we review this equivalence and verify that the folklore profunctor optic laws are equivalent to lawfulness as defined here.

@@

@@

  Note that because $\iota(S, S') = (S, S')$, required equations for $\iota$ to be a monoidal functor hold by definition (although we don't yet know that $\Optic_\C$ is monoidal). The pentagon and triangle equations then hold in $\Optic_\C$, as they are the image of the same diagrams in $\C \times \C^\op$ under $\iota$. The only remaining thing to verify is that these structure maps are natural in $\Optic_\C$, but this follows from the previous proposition.
\end{proof}

@@

@@

@@

\begin{definition}[Compare {\cite[Definition 5.1]{CoherenceForLenses}}]
  A \emph{teleological category is} a symmetric monoidal category $(\T, \teletimes, I)$, equipped with:
  \begin{itemize}
  \item A symmetric monoidal subcategory $\T_d$ of \emph{dualisable morphisms} containing all the objects of $\T$, with an involutive symmetric monoidal functor ${(-)}^* : \T_d \to \T_d^\op$, where---not finding a standard symbol for such a thing---we mean $\T_d^\op$ to be the category with both the direction of the arrows \emph{and} the order of the tensor flipped: ${(A \otimes B)}^* \cong B^* \otimes A^*$. Note that there is therefore also a canonical isomorphism $\phi : I \cong I^*$
  \item A symmetric monoidal extranatural family of morphisms $\varepsilon_X : X \otimes X^* \to I$, called \emph{counits}, natural with respect to the dualisable morphisms.
  \end{itemize}
\end{definition}
Unpacking the definition, $\varepsilon$ being a symmetric monoidal extranatural transformation amounts to the following diagrams in $\T$ commuting:
\[
  \begin{tikzcd}
    X \teletimes Y^* \ar[r, "f \teletimes Y^*"]  \ar[d, "X \teletimes f^*", swap] & Y \teletimes Y^* \ar[d, "\varepsilon_Y"] \\
    X \teletimes X^* \ar[r, "\varepsilon_X", swap] & I
  \end{tikzcd} \hspace{1cm}
  \begin{tikzcd}
    X^* \teletimes X \ar[r, "s"]  \ar[dr, "\varepsilon_{X^*}", swap] & X \teletimes X^* \ar[d, "\varepsilon_X"] \\
    & I
  \end{tikzcd}\]
\[
  \begin{tikzcd}[column sep = large]
    X \teletimes Y \teletimes Y^* \teletimes X^* \ar[r, "X \teletimes \varepsilon_Y \teletimes X^*"]  \ar[dr, "\varepsilon_{X \teletimes Y}", swap] & X \teletimes X^* \ar[d, "\varepsilon_X"] \\
    & I
  \end{tikzcd} \hspace{1cm}
  \begin{tikzcd}
    I \teletimes I^* \ar[r,"I \teletimes \phi"] \ar[dr, swap, "\varepsilon_I"] & I \ar[d, equal] \\
    & I
  \end{tikzcd}
\]
where $f : X \to Y$ is dualisable.

@@

@@

@@

@@

@@

In this section we make a different definition of lawful that at first seems strange, but which in the case of lenses corresponds \emph{exactly} to the three concrete lens laws with no additional assumptions on $\C$ required. As further justification for this definition, in Section~\ref{sec:profunctor-optics} we will see an interpretation of (unlawful) optics as maps between certain comonoid objects. Lawfulness in our sense corresponds exactly to this map being a comonoid homomorphism.

@@

\begin{proposition}\label{prop:lawful-lens-laws}
  A concrete lens given by $\fget$ and $\fput$ is lawful (in our sense) iff it obeys the three concrete lens laws.
\end{proposition}
\begin{proof}
  We begin by giving $\Twoptic_\times(S, A)$ the same treatment as $\Optic_\times(S, A)$. Using Yoneda reduction and the universal property of the product as we did for lenses themselves, we have:
  \begin{align*}
  \Twoptic_\times(S, A)
  &= \int^{M_1, M_2 \in \C} \C(S, M_1 \times A) \times \C(M_1 \times A, M_2 \times A) \times \C(M_2 \times A, S) \\
  &\cong \int^{M_1, M_2 \in \C} \C(S, M_1) \times \C(S, A) \times \C(M_1 \times A, M_2 \times A) \times \C(M_2 \times A, S) \\
  &\cong \int^{M_2 \in \C} \C(S, A) \times \C(S \times A, M_2 \times A) \times \C(M_2 \times A, S) \\
  &\cong \int^{M_2 \in \C} \C(S, A) \times \C(S \times A, M_2) \times \C(S \times A, A) \times \C(M_2 \times A, S) \\
  &\cong \C(S, A) \times \C(S \times A, A) \times \C(S \times A \times A, S)
  \end{align*}
%  Call this last set $\conctwice_\times(S, A)$. Written equationally, the isomorphism $\Phi : \Twoptic_\times(S, A) \to \conctwice_\times(S, A)$ is given by:
  Written equationally, the isomorphism $\Phi : \Twoptic_\times(S, A) \to \C(S, A) \times \C(S \times A, A) \times \C(S \times A \times A, S)$ is given by:
  \begin{align*}
    \Phi(\repthree{l}{c}{r}) = (&\pi_2 l, \\
                                   &\pi_2 c (\pi_1 l \times A), \\
                                   &r(\pi_1 c (\pi_1 l \times A) \times A))
  \end{align*}
  Now suppose we are given a lens $p$ that corresponds concretely to $(\fget, \fput)$, so $p = \rep{[\id_S, \fget]}{\fput}$. Evaluating $\outside$ on this gives:
  \begin{align*}
    \outside(\rep{[\id_S, \fget]}{\fput}) = \fput [\id_S, \fget]
  \end{align*}
  so requiring $\outside(p) = \id_S$ is precisely the $\fget\fput$ law.

@@

This characterisation of setters is maybe a little odd, in that we have ended up with a function of $\Set$s, rather than a description internal to $\C$. If we modify our definition of $\Setter$, we do get an internal characterisation. Suppose $\C$ is cartesian closed and let $\Strong_\C$ be the category of \emph{strong functors} on $\C$.

\begin{definition}[\cite{StrongFunctors}]\label{def:strong-functor}
  A \emph{(left) strong functor} is a functor $F : \C \to \C$ equipped with a natural transformation called the \emph{strength}:
  \begin{align*}
    \theta_{A,B} : A \otimes F B \to F(A \otimes B)
  \end{align*}
  such that the strength commutes with the unitor:
  \[
    \begin{tikzcd}
      I \otimes F A \ar[r, "\theta_{1,A}"] \ar[d, "\cong" left]  & F(I \otimes A) \ar[d, "\cong" right] \\
      F A \ar[r, equals] & F A
    \end{tikzcd}
  \]
  and with associativity:
  \[
    \begin{tikzcd}
      (A \otimes B) \otimes F C \ar[rr, "\theta_{A \otimes B, C}"] \ar[d, "\alpha_{A,B,FC}" left]  && F((A \otimes B) \otimes C) \ar[d, "F\alpha_{A,B,FC}" right] \\
      A \otimes (B \otimes F C) \ar[r, "A \otimes \theta_{B,C}" below] & A \otimes F(B \otimes C) \ar[r, "\theta_{A, B\otimes C}" below] & F(A \otimes (B \otimes C))
    \end{tikzcd}
  \]
  A \emph{strong natural transformation} $\tau : (F,\theta) \Rightarrow (G,\theta')$ is a natural transformation that respects the strengths. There is an evident category $\Strong(\C)$ of strong endofunctors and strong natural transformations, and a forgetful functor $U : \Strong(\C) \to [\C, \C]$.
\end{definition}
Then, again, $\Strong_\C$ acts on $\C$ by evaluation. We leave it to the reader to verify there is a natural isomorphism \[\C(S, FA) \cong \Strong_\C(\homC(A, -) \times S, F),\] which we can use to describe optics for this action as elements of $\C(\homC(A, A'), \homC(S,S'))$.
%The use of strong functors here is due to the correspondence between tensorial strengths and ``$\C$-enrichments'' on a functor.

@@

@@

\item \emph{Grates}~\cite{GratesPost} are optics for the contravariant action of a monoidal closed category $\C$ on itself by $X \cdot A \mapsto \homC(X, A)$. Concretely these correspond to morphisms \[ \C(\homC(\homC(S, A), A'), S'). \]
\end{itemize}
